---
title: "Crime and COVID-19 in Houston: How do they relate?"
author: "Gail Oudekerk, Kelly Zeng, Elijah Sales, Naghmeh Hosseini, Jennifer Jia"
date: "4/21/2022"
output: ioslides_presentation
runtime: shiny
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Research Question

Is there a relationship between COVID-19 positivity rates and crime offense rates in Houston, Texas? 

## Datasets

- Crime by street/neighborhood in Houston from 2017 to 2021
  - Incident, OccurrenceDate, OccurenceHour, NIBRSClass, NIBRSDescription, OffenseCount, Beat, Premise, StreetNo, BlockRange, StreetName, StreetType, Suffix, City, ZIPCode

- Daily COVID-19 positivity rate in Harris County from 2020 to 2022
  - ï..OBJECTID, id, Date, Positive_Pct, Postive_Pct_14dayAvg, Threshold_5PCT, Status, Red_Status, Orange_Status, Yellow_Status, Green_Status, Total, Negative, Positive, Date_Str, Positive_Pct_Str, Postive_Pct_14dayAvg_Str, Threshold_5PCT_Str, Date_Updated, today

- News headlines mentioning COVID-19

## Initial Visualizations

- Plot 1

```{r}
library(ggplot2)
setwd ("C:\\Users\\Kelly\\Downloads\\2021Sem2\\STAT405\\STAT405Project\\Data")
load(file = "crimedata.Rda")
# Pie Chart of Types of Crimes
ggplot(data = crimedata) + aes(x = factor(1), fill = NIBRSDescription) + geom_bar(stat = "count") + 
  theme(title = element_text(size = 17)) + 
  labs(title = "Number of Offenses by Type of Crime") +
  coord_polar(theta = 'y', start = 0) + theme_void() + 
  theme(plot.title = element_text(hjust = 0.5)) 
```

## Initial Visualizations

- Plot 2
```{r}
setwd ("C:\\Users\\Kelly\\Downloads\\2021Sem2\\STAT405\\STAT405Project\\Data")
load(file = "covid.rda")
library(ggplot2)
covid1 <- tibble::rowid_to_column(covid, "ID")
ggplot(covid1)+aes(x=id, y=Positive)+labs(y = "Positive Covid Cases", x = "Days after 2020/05/01", title = "Positive Covid Cases by Day")+geom_line()+theme(plot.title = element_text(hjust = 0.5))
```

## Initial Visualizations

- Plot 3
```{r}
crimedata1 <- tibble::rowid_to_column(crimedata, "ID")
crimecomp <- crimedata1[crimedata1$ID>= 611320, ]
ggplot(crimecomp)+aes(x=OccurrenceDate)+geom_line(stat = "Count")+ xlab("Days after 2020/05/01") + ylab("Number of Offenses") 
```


## Question 1
- Is there a correlation between COVID cases and crime cases?
- Initial hypothesis is that there are more crimes on days with worse COVID alert levels

## Question 1

```{r}
library(ggplot2)
suppressWarnings(library(ggpubr))
suppressMessages(library(dplyr))
suppressMessages(library("gridExtra"))

# Group by NIBRSDescription

# Changing the date format in the Harris County COVID Dataset
covid$OccurrenceDate <- strptime(covid$Date_Str, "%Y/%m/%d")

covid$OccurrenceDate <- as.Date(covid$OccurrenceDate)
merged_data <- merge(covid, crimedata)

# Subsetting data based on status types
red_status <- merged_data[merged_data$Red_Status == 1, ]
orange_status <- merged_data[merged_data$Orange_Status == 1, ]
yellow_status <- merged_data[merged_data$Yellow_Status == 1, ]
green_status <- merged_data[merged_data$Green_Status == 1, ]

cbPalette <- c("#999999", "#E69F00", "#56B4E9", "#009E73",
               "#F0E442", "#0072B2", "#D55E00", "#CC79A7")

# Red status
prod <- ggplot(data = red_status) + aes(x=NIBRSDescription)
red_bar_plot <- prod + geom_bar(data = red_status, fill = "red") + theme(axis.text.x = element_text(angle = 35, hjust=1, size = 5))+ ggtitle("Crimes during Red Status days") + scale_fill_manual(values = cbPalette)

# Orange status
prod <- ggplot(data = orange_status) + aes(x=NIBRSDescription)
orange_bar_plot <- prod + geom_bar(data = orange_status, fill = "orange") + theme(axis.text.x = element_text(angle = 35, hjust=1, size = 5)) + ggtitle("Crimes during Orange Status days") + scale_fill_manual(values = cbPalette)

# Yellow status
prod <- ggplot(data = yellow_status) + aes(x=NIBRSDescription)
yellow_bar_plot <- prod + geom_bar(data = yellow_status, fill = "yellow") + theme(axis.text.x = element_text(angle = 35, hjust=1, size = 5)) + ggtitle("Crimes during Yellow Status days") + scale_fill_manual(values = cbPalette)

# Green status
prod <- ggplot(data = green_status) + aes(x=NIBRSDescription)
green_bar_plot <- prod + geom_bar(data = green_status, fill = "green") + theme(axis.text.x = element_text(angle = 35, hjust=1, size = 5)) + ggtitle("Crimes during Green Status days") + scale_fill_manual(values = cbPalette)


grid.arrange(red_bar_plot, orange_bar_plot, yellow_bar_plot, green_bar_plot, ncol = 2)
```

## Question 2

- Why are there more occurrences of certain types of crimes compared to others?
- Our inital hypothesis was that theft & robbery increased a lot due to COVID-19
- 2019->2020, theft & robbery decreased
- 2020-2021, theft & robbery remained constant
- Contrary to our initial hypothesis

## Question 2
```{r}
# filter(crimedata1, str_detect(crimedata1$OccurrenceDate, "2021-12-31"))

#crime17 <- crimedata1[crimedata1$ID>= 1 & crimedata1$ID <= 118582, ]
#ggplot(data = crime17) + aes(x = NIBRSDescription) + geom_bar(stat = "count", fill = c("#AED1EF", "#C1D5E2", "#D4D9D5", "#E8DDC7", "#F2DFC1")) + 
  #theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 9), 
   #     axis.text.y = element_text(size = 9), 
    #    axis.title = element_text(size = 11), 
     #   title = element_text(size = 13)) + 
#  xlab("Type of Crime") + ylab("Number of Offenses") + 
 # labs(title = "Number of Offenses by Type of Crime 2017") +
#  theme(plot.title = element_text(hjust = 0.5)) 

crime18 <- crimedata1[crimedata1$ID>= 118582 & crimedata1$ID <= 311246, ]
plot18 <- ggplot(data = crime18) + aes(x = NIBRSDescription) + geom_bar(stat = "count", fill = c("#AED1EF", "#C1D5E2", "#D4D9D5", "#E8DDC7", "#F2DFC1", "#F2D9C8", "#F1CED5", "#F0B9EF")) + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 6), 
        axis.text.y = element_text(size = 4), 
        axis.title = element_text(size = 7), 
        title = element_text(size = 10)) + 
  xlab("Type of Crime") + ylab("Number of Offenses") + 
  labs(title = "Number of Offenses by Type of Crime 2018") +
  theme(plot.title = element_text(hjust = 0.5)) 

crime19 <- crimedata1[crimedata1$ID>= 311246 & crimedata1$ID <= 530280, ]
plot19 <- ggplot(data = crime19) + aes(x = NIBRSDescription) + geom_bar(stat = "count", fill = c("#AED1EF", "#C1D5E2", "#D4D9D5", "#E8DDC7", "#F2DFC1", "#F2D9C8", "#F1CED5")) + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 6), 
        axis.text.y = element_text(size = 4), 
        axis.title = element_text(size = 7), 
        title = element_text(size = 10)) + 
  xlab("Type of Crime") + ylab("Number of Offenses") + 
  labs(title = "Number of Offenses by Type of Crime 2019") +
  theme(plot.title = element_text(hjust = 0.5)) 

crime20 <- crimedata1[crimedata1$ID>= 530280 & crimedata1$ID <= 778191, ]
plot20 <- ggplot(data = crime20) + aes(x = NIBRSDescription) + geom_bar(stat = "count", fill = c("#AED1EF", "#C1D5E2", "#D4D9D5", "#E8DDC7", "#F2DFC1", "#F2D9C8", "#F1CED5", "#F0B9EF")) + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 6), 
        axis.text.y = element_text(size = 4), 
        axis.title = element_text(size = 7), 
        title = element_text(size = 10)) + 
  xlab("Type of Crime") + ylab("Number of Offenses") + 
  labs(title = "Number of Offenses by Type of Crime 2020") +
  theme(plot.title = element_text(hjust = 0.5)) 

crime21 <- crimedata1[crimedata1$ID>= 778191 & crimedata1$ID <= 1025169, ]
plot21 <- ggplot(data = crime21) + aes(x = NIBRSDescription) + geom_bar(stat = "count", fill = c("#AED1EF", "#C1D5E2", "#D4D9D5", "#E8DDC7", "#F2DFC1", "#F2D9C8", "#F1CED5", "#F0B9EF")) + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 6), 
        axis.text.y = element_text(size = 4), 
        axis.title = element_text(size = 7), 
        title = element_text(size = 10)) + 
  xlab("Type of Crime") + ylab("Number of Offenses") + 
  labs(title = "Number of Offenses by Type of Crime 2021") +
  theme(plot.title = element_text(hjust = 0.5)) 
# for some reason, this part shows up really awkwardly on my console, not sure if shiny helps

grid.arrange(plot18, plot19, plot20, plot21, ncol = 2)
```


## Question 3
- Does crime in Houston, Texas change by time of day in response to COVID-19?
- The initial hypothesis was that COVID-19 did change the trend of Houston crime data over time

## Question 3
```{r}
##Houston Crime by Time of Day(2017-2021) USING ggplot2
library(ggplot2)
setwd ("C:\\Users\\Kelly\\Downloads\\2021Sem2\\STAT405\\STAT405Project\\Data")
load(file = "crimedata.Rda")
# Convert hour column from character to numeric"
OccurrenceHour2<- as.numeric(crimedata$OccurrenceHour)
crimedata2 <- cbind(crimedata,OccurrenceHour2)
crimedata2['OccurrenceHour2'][crimedata2['OccurrenceHour2'] == 24] <-  0

count <- table(crimedata2$NIBRSDescription, crimedata2$OccurrenceHour2)


cbPalette <- c("#999999", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
theme1 <- theme(axis.text = element_text(size = 8, colour = "#6b3447"),
          axis.title = element_text(size = 10, colour = "#2f2f63"),
          legend.text = element_text(size = 8, colour = "#6b3447"),
          legend.position = "top",
          title = element_text(size = 13, colour = "#2f2f63"),
          axis.ticks = element_line(colour = "#6b3447"),
          axis.text.x = element_text(angle = 45, hjust = 1, size = 10),
          axis.text.y = element_text(angle = 0, hjust = 1, size = 10),
          axis.title.x = element_text(size = 11),
          axis.title.y = element_text(size = 11))

ggplot(data = crimedata2) + aes(x =OccurrenceHour2, fill=NIBRSDescription) + geom_bar(stat = "Count") + xlab("Hour") + ylab("Count") + scale_x_continuous(breaks=seq (0,23,by=1)) + scale_y_continuous(breaks=seq (0,70000,by=10000)) + labs(title = "Houston Crime by Time of Day(2017-2021)")+theme(plot.title = element_text(hjust=0.5)) + labs(fill='') + theme1 + scale_fill_manual(values=cbPalette) 

```

## Question 3
```{r}
library(ggplot2)
# Houston Crime by Time of Day(2017-2019) to see correlation
crimedata3 <- crimedata2[crimedata2$OccurrenceDate >= "2017-01-01" & crimedata2$OccurrenceDate <= "2019-12-31", ]

count <- table(crimedata3$NIBRSDescription, crimedata3$OccurrenceHour2)

         cbPalette <- c("#999999", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
         theme1 <- theme(axis.text = element_text(size = 8, colour = "#6b3447"),
                 axis.title = element_text(size = 10, colour = "#2f2f63"),
                 legend.text = element_text(size = 8, colour = "#6b3447"),
                 #legend.position = "top",
                 title = element_text(size = 13, colour = "#2f2f63"),
                 axis.ticks = element_line(colour = "#6b3447"),
                 axis.text.x = element_text(angle = 45, hjust = 1, size = 10),
                 axis.text.y = element_text(angle = 0, hjust = 1, size = 10),
                 axis.title.x = element_text(size = 11),
                 axis.title.y = element_text(size = 11))

plot2 <- ggplot(data = crimedata3) + aes(x =OccurrenceHour2, fill=NIBRSDescription) + geom_bar(stat = "Count") + xlab("Hour") + ylab("Count") + scale_x_continuous(breaks=seq (0,23,by=1))+scale_y_continuous(breaks=seq (0,35000, by=10000))+labs(title = "Houston Crime by Time of Day(2017-2019)")+theme(plot.title = element_text(hjust=0.5))+labs(fill='')+theme1+scale_fill_manual(values=cbPalette)



crimedata4 <- crimedata2[crimedata2$OccurrenceDate >= "2020-01-01" & crimedata2$OccurrenceDate <= "2021-12-31", ]
  
count <- table(crimedata3$NIBRSDescription, crimedata3$OccurrenceHour2)

         cbPalette <- c("#999999", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
         theme1 <- theme(axis.text = element_text(size = 6, colour = "#6b3447"),
                 axis.title = element_text(size = 10, colour = "#2f2f63"),
                 legend.text = element_text(size = 8, colour = "#6b3447"),
                 #legend.position = "top",
                 title = element_text(size = 13, colour = "#2f2f63"),
                 axis.ticks = element_line(colour = "#6b3447"),
                 axis.text.x = element_text(angle = 45, hjust = 1, size = 10),
                 axis.text.y = element_text(angle = 0, hjust = 1, size = 10),
                 axis.title.x = element_text(size = 11),
                 axis.title.y = element_text(size = 11))

plot3 <- ggplot(data = crimedata4) + aes(x =OccurrenceHour2, fill=NIBRSDescription) + geom_bar(stat = "Count") + xlab("Hour") + ylab("Count") + scale_x_continuous(breaks=seq (0,23,by=1))+ labs(title = "Houston Crime by Time of Day(2020-2021)")+theme(plot.title = element_text(hjust=0.5))+labs(fill='')+ theme1+scale_fill_manual(values=cbPalette)+ scale_y_continuous (breaks=seq (0,35000, by=10000))

grid.arrange(plot2, plot3, nrow = 2)
```

## Killer Plot
```{r}
suppressWarnings(library(RSQLite))
suppressWarnings(library(stringr))
suppressWarnings(library(grid))
suppressWarnings(library(shiny))

setwd ("C:\\Users\\Kelly\\Downloads\\2021Sem2\\STAT405\\STAT405Project\\Data")

#create connection
dcon <- dbConnect(SQLite(), dbname = "STAT405_crimeCOVIDfinal.sqlite")

invisible(suppressMessages(dbSendQuery(conn = dcon, "
PRAGMA foreign_keys = ON;
")))

query <- paste0("
 SELECT * from (select OccurrenceDate, NIBRSDescription, count(NIBRSDescription) as total
    FROM crimedata
    GROUP BY OccurrenceDate, NIBRSDescription) AS c
    INNER JOIN (SELECT Date_Format AS OccurrenceDate, Positive_Pct, COVID_daily_status
                  FROM covid) AS o
                ON c.OccurrenceDate = o.OccurrenceDate")
suppressWarnings(res <- dbSendQuery(dcon, query))
killerdata<- dbFetch(res, -1)[,-c(4)]
dbClearResult(res)

dbDisconnect(dcon)

#abbreviate alert status to only color name (more aesthetically pleasing colors)
killerdata$COVID_daily_status <- str_replace(killerdata$COVID_daily_status, "_Status", "")
killerdata$COVID_daily_status <- str_replace(killerdata$COVID_daily_status, "Red", "firebrick3")
killerdata$COVID_daily_status <- str_replace(killerdata$COVID_daily_status, "Orange", "darkorange2")
killerdata$COVID_daily_status <- str_replace(killerdata$COVID_daily_status, "Yellow", "gold1")
killerdata$COVID_daily_status <- str_replace(killerdata$COVID_daily_status, "Green", "palegreen3")

makeplot <- function(data, date_str){
  #data = killerdata
  #date_str = "2020-05-01"
  grid.newpage()
  vp <- viewport(x = 0.5, y = 0.5, 
                 width = 0.7, height = 0.7, 
                 just = c("center", "center"),
                 xscale = c(0,1), yscale = c(0, 100))
  #grid.show.viewport(vp)
  pushViewport(vp)
  #grid.rect(gp = gpar(col = "red"))
  
  day <- data[data$OccurrenceDate == date_str,]
  color <- as.character(day$COVID_daily_status[1])
  #print(color)
  pos_rate <- day$Positive_Pct[1]
  
  grid.rect(x = 0.001, y = 0,
            width = 0.999, height = pos_rate / 100,
            gp = gpar(col = color, fill = color),
            just = c("left", "bottom"))
  #grid.text("Origin", x = 0, y = 0, just = c("left", "bottom"))
  #grid.yaxis(at = c(0, pos_rate, seq(20, 100, by = 20)))
  grid.yaxis(at = c(0, pos_rate, 100),
             gp = gpar(fontsize = 12))
  
  bar_vp <- viewport(x = 0.5, y = pos_rate / 100 + 0.18,
                     width = 0.8, height = 0.57, 
                     just = c("center", "bottom"),
                     yscale = c(0, 310))
  pushViewport(bar_vp)
  #grid.rect(gp = gpar(col = "red"))
  grid.yaxis(main = FALSE,
             gp = gpar(col = "gray48", fontsize = 12))
  
  #create bars
  num_bars <- dim(day)[1]
  spacing <- 1 / num_bars
  spacing_vect <- seq(0 + (spacing / 8), (1-spacing/8), by = spacing)
  
  for(i in 1:num_bars){
    grid.rect(x = spacing_vect[i], y = 0,
              width = 2 * spacing / 3, height = (day$total[i] / 310),
              gp = gpar(col = "gray30", fill = "gray30"),
              just = c("left", "bottom"))
    grid.text(label = day$NIBRSDescription[i],
              x = spacing_vect[i], y = (day$total[i] / 310 + 0.04),
              just = c("left", "top"), rot = 75,
              gp = gpar(fontsize = 12, col = "gray48"))
    
    stack_vp <- viewport(x = spacing_vect[i], y = day$total[i] / 310,
                         width = 2 * spacing / 3, height = 0.08, 
                         just = c("left", "top"),
                         clip = "on")
    pushViewport(stack_vp)
    grid.circle(x = 0.5, y = 1.75, r=1.5,
                gp = gpar(col = "gray30", fill = "gray48"))
    popViewport()
  }
  
  popViewport()
  
  day_total <- sum(day$total)
  
  inner_vp <- viewport(x = 0.5, y = pos_rate / 100,
                       width = 1, height = 0.18, 
                       just = c("center", "bottom"),
                       clip = "on")
  pushViewport(inner_vp)
  grid.circle(x = 0.5, y = 3.5, r=4, 
              gp = gpar(col = "gray30", fill = "gray30"))
  grid.text(paste0("Day's Observed Crime Count = ", day_total), 
            x =0.5, y = 0.6, 
            just = c("center", "center"),
            gp = gpar(col = "white", fontsize = 16))
  popViewport()
  
  wake_vp <- viewport(x = 0.78, y = pos_rate / 100,
                       width = 0.15, height = 0.05, 
                       just = c("center", "center"),
                       clip = "on")
  pushViewport(wake_vp)
  print("now")
  grid.circle(x = 0.5, y = -1.2, r=2.2, 
              gp = gpar(col = color, fill = color))
  #grid.rect(gp = gpar(col = "green"))
  popViewport()
  
  popViewport()
  
  label_vp <- viewport(x = 0.5, y = 0.5, 
                 width = 0.9, height = 0.9, 
                 just = c("center", "center"))
  pushViewport(label_vp)
  grid.text(label = paste0("Houston Daily Crime by Type (with COVID-19 levels)"),
            x = 0.5, y = 0.07,
            just = c("centre"),
            gp = gpar(fontsize = 24, col = "black"))
  grid.text(label = paste0("color = COVID alert level"),
            x = 0.75, y = 0,
            just = c("centre"),
            gp = gpar(col = ifelse(color == "gold1", "black", color),
                      fontsize = 15))
  grid.text(label = paste0("Logged Crime Count (by Houston Police)"),
            x = 0.93, y = 0.5,
            rot = 90,
            just = c("centre"),
            gp = gpar(col = "gray48", fontsize = 15))
  grid.text(label = paste0("Houston Positive Test Rate (COVID-19)"),
            x = 0, y = 0.5,
            rot = 90,
            just = c("centre"),
            gp = gpar(col = "black", fontsize = 15))
  popViewport()
}

inputPanel(
  dateInput("dateID", label = "Date: ",
           value = "2020-05-01", min = "2020-05-01", max = "2021-12-31",
           format = "yyyy-mm-dd")
)

renderPlot(makeplot(killerdata, input$dateID), width = 594, height = 432)
```


## Conclusion
- Crimes across all offense types appear more frequent with worsening COVID alert levels
- Even though COVID cases continued to increase over the years, the most common crime cases remained stable
- COVID-19 changed the time of occurrence in some violations, but the general trend remained the same
- Building customizable, flexible plots in grid can help us convey complex data trends in original, interesting ways
